A Component-Based Sensor Network for \\ Environmental Monitoring

### 2009-09-29 docs/icccn2009/paper.tex
These are reviews and updates to the text before the camera-ready.
I've updated paragraphs that repeats the same contents in order
to be back to 6 pages.

git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@606 559edde2-a4d4-44e0-96fc-4035e40d7396

\author{\authorblockN{A. Puder, T. Johnson, K. Sales, M. de Sales}
\authorblockA{San Francisco State University \\
Computer Science Department \\
1600 Holloway Avenue \\
San Francisco, CA 94132 \\
EMail: \{arno$\mid$tlj$\mid$kleber$\mid$msales\}@sfsu.edu}
\and
\authorblockN{D. Robinson}
\authorblockA{San Francisco State University \\
Romberg Tiburon Center \\
3150 Paradise Drive \\
Tiburon, CA 94920 \\
EMail: dhr@sfsu.edu}}

\maketitle
\begin{abstract}
  In this paper we describe a sensor network for environmental
  monitoring based on highly specialized, off-the-shelf sensing
  devices.  Environmental monitoring depends on the reliable
  collection of measurements from remotely deployed sensors and the
  rapid transfer of those measurements to data centers. Typically,
  sensors are deployed in out-of-the-way locations, where physical
  access is limited and connections for telemetry are poor or
  nonexistent. Maintaining the data stream from these sensors places
  demands on human resources, requiring site visits to service the
  sensors and to download data stored on internal memory.  The work
  described in this paper is a collaborative effort between the
  Computer Science Department at San Francisco State University and
  the Romberg Tiburon Center. The result of this collaboration is an
  environmental sensor network deployed in the San Francisco Bay. It
  provides means to program and interrogate sensors at the remote
  field locations and to transmit collected measurements to a remote
  database.
\end{abstract}

@@

The work described in this paper is a collaborative effort between the
Computer Science Department at San Francisco State University and the
Romberg Tiburon Center (RTC), a research institute focusing on the
understanding of complex marine and estuarine environments. The result
of this collaboration is NetBEAMS (Networked Bay Environmental
Assessment Monitoring System), an environmental sensor network
deployed in the San Francisco Bay. NetBEAMS provides the means to program
and interrogate sensors at remote field locations and to transmit
collected measurements to a database, where the data is rapidly
processed, archived, and made available to potential users in near
real-time. In the following sections we give an overview of the NetBEAMS
architecture and implementation.  Section \ref{SEC_BACKGROUND}
describes a typical environmental sensor used by the RTC. Section
\ref{SEC_DSP} describes our Data Sensor Platform (DSP) that allows us
to efficiently build up an end-to-end environmental sensor network while
Section \ref{SEC_NETBEAMS} describes NetBEAMS. Finally, Section 
\ref{SEC_CONCLUSION} presents our conclusions and future works.

@@

Environmental researchers typically use standard, off-the-shelf
sensing devices for their purposes. These sensing devices are at the
opposite end of the size spectrum of miniature sensors such as SmartDust. One such
device is the YSI 6600EDS V2 Sonde \cite{Sonde01}, a water
quality monitoring device that gathers water quality data, and
operates in fresh, sea, or polluted water. Some of the measurements
that the YSI is capable of are conductivity, temperature, chloride,
ammonium, nitrate, turbidity, and chlorophyll.  It is under 55 cm in
length, 8.9 cm in diameter, and weighs approximately 3.18 kg. It
operates at temperatures between -5 to +50 deg C and at depths up to
200 meters.  It uses 8 C-size Alkaline Batteries or External 12 VDC.
Battery life can last up to 75 days depending on sensor configuration.
The Sonde's bulkhead contains multiple pin connectors to support
sensor probes for measuring parameters as shown in Table
\ref{TAB_SONDE_MEASUREMENTS}.

@@

Each probe may have multiple sensors.  A computer can interface with
the Sonde via RS-232C or SDI-12. The Sonde can be configured to
collect data samples in discrete or unnattended mode.  Discrete mode
is normally performed while a technician manages the Sonde during the
sampling process. In this mode, the Sonde is connected via 650
MDS Display/Logger or a serial cable connected to a PC, where the sampling
frequency is usually set to a high frequency. In contrast, Unattended
mode is performed when the Sonde is deployed offshore or in a
remote body of water, where the sampling frequency is usually set for a
longer period of time (i.e. 5 - 15 min). Assuming a 5 min sampling
frequency, the Sonde produces 267 KB of data per day.
In order to be configured,
the Sonde offers a set of menus that can be accessed through a terminal session
over RS232. In this way, the Sonde can be configured to
start logging data to an internal file. In order to view the data as it's
being captured, the option to show live data is chosen and the data can then be 
seen via a serial terminal session at each interval.

We provide these details to show that sensing devices such as the YSI
are highly specialized devices with a unit price of several thousand
dollars. These sensors are usually sparsely deployed in remote
locations that not only make their maintenance difficult, but also
access to their measurements difficult.  Typically, the sensors are autonomous
in the sense that they can run for a certain amount of time on battery power and
store the measurement data in internal memory. Only when the sensors
are serviced, the measurement data, which was collected since the last
servicing, is uploaded to a portable storage device. From there, the
data can be uploaded to server. This manual procedure to retrieve
sensor data is not only error prone, but the results greatly differ
between the time the sensor performs the measurement and when
it becomes available for further processing. For many applications,
however, it would be beneficial to have near-realtime access to the
sensor measurements. We identify the following requirements that an
end-to-end infrastructure supporting devices such as the YSI should
offer:

@@

Sensor networks, especially environmental sensors, are normally
deployed over a geographically wide area, where the communication
infrastructure is not always reliable or, at least, not coutinuously
available. As the sensor network grows in size and complexity, 
scalability, reliability and maintainability become critical issues.
For instance, additional components might be added to the network, sensor
configuration might be altered, or new versions of components might have to be
deployed. As a consequence, the sensor network must be able to support such
adjustments without significant overhead for maintainers.  From a system
development and integration point of view, interfaces among components
and between components and the network should be simple. As a result, complex
interaction can be constructed on top of these simple interfaces. In
the following subsections, we discuss how our proposed architecture addresses
those issues.

@@

NetBEAMS's architecture is based on a micro-kernel approach. The core
functionality is limited to a minimum and all extra services are built
as plug-and-play modules on top of the micro-kernel. With the help of
this approach, we can keep the basic infrastructure compact while
allowing customization via special-purpose modules. In our terminology,
we call the basic foundation the \emph{Data Sensor Platform}
(DSP). DSP is collection of local built-in and plug-in components; 
a sensor network is built from several DSPs that are linked
with each other.

Figure \ref{FIG_DSP} depicts the basic architecture of the DSP. We
refer to the plug-and-play modules as DSP Components. DSP Components
are self-contained modules that can be added and removed to a DSP
installation at runtime. The basic communication unit within a DSP
is a message. DSP Components can exchange messages of different types.
Actually, all interactions between DSP Components are implemented
through messages yielding a simple but powerful communication
structure. Moreover, we distinguish between \emph{Data Producers} (DP) and
\emph{Data Consumers} (DC) within the scope of a DSP. DP and DC can be
seen as roles that a DSP Component may have, as it may also 
act in the role of DC and DP at the same time. Messages
in DSPs are routed based on rules, which can be configured statically
or dynamically. Rules for routing messages can be based on different
information such as producers or consumers of the message, the message
itself, or specific configuration parameters provided externally.

@@

The DSP is responsible for routing messages between DPs and DCs. It is
important to note that this routing only happens within a
DSP. Following the paradigm of a micro-kernel approach, the DSP has no
notion of remote communication or sensor-specific details. All this
knowledge is embodied in special purpose DSP Components.

@@

A DSP Component encapsulates a particular functionality. We
distinguish between those components that produce data (DP) and those
that consume data (DC). An example for a DSP Component acting in the
role of a DP is a module that interfaces with a physical sensor. Such
a module reads data from the sensor and converts the data into a
message that is forwarded to the DSP. Because this module generates
data with respect to the DSP, it is a data producer. An example for a
data consumer is a converter module that converts the internal,
DSP-specific data format to a format suitable for consumption by
external applications. One such common format used in the domain of
environmental monitoring is OpenDAP \cite{opendap01}.  Since the
converter module accepts messages from the DSP, it acts in the role of
a data consumer. Finally, a DSP Component implementing persistence is
an example of a module that acts in the roles of both data producer
and consumer, since it accepts messages from the DSP and stores them in a
database. Furthermore, it may act as a data producer by resending previously
stored messages at a later time.

@@

By implementing the above Java interface, the resulting code
effectively becomes a DSP Component that can be deployed on a DSP at
runtime. A component is identified by its type (line 3). Only one
component of a certain type per DSP is permissible. A type identifier
must be unique across all DSP Components in a DSP. Upon deploying a
component, an initialization function is called (line 4). Once a
component is initialized, it may be started (line 6) and stopped (line
7) multiple times. Whenever the DSP wishes to pass a message (that was
generated by some DP), it invokes the \texttt{deliver()} method (line
8). A DP can send messages via a \texttt{DSP Broker} (a DSP Component itself) 
obtained from \texttt{DSPContext} object that was
passed to the DP during initialization.

@@

The broker can deliver messages concurrently and asynchronously. It
sets any additional message header information necessary for proper
routing. The broker does not make any routing decisions. It delegates
this responsibility to the \emph{matcher}. The matcher is responsible
for matching messages sent by a DP and routes it to one or more DCs.
This decision is based on a set of routing rules (see Figure
\ref{FIG_MATCHER_RULE}) which can be statically or dynamically
configured.

@@

The matcher's main purpose is to achieve a loose coupling between
DCs and DPs. DPs may create messages without specifying a specific
destination DC. Once the message is sent through the DSP, the matcher will
determine which DCs should receive the message. Note that, a DP can
still specify a DC for its message. In this case, the matcher includes
this DC in a list of DCs to receive the message.

@@

The Gumstix runs Linux 2.6 with the BusyBox utilities, and uses the
OpenEmbedded cross-compile build environment to provide a complete
Linux environment and a large range of Linux applications for the ARM
architecture.  We use the JamVM \cite{jamvm01} implementation of Sun
Microsystem's virtual machine in order to run the DSP. The Gumstix
communicates with the YSI Sonde via an RS232 serial port using a
null-modem cable, since both the Sonde and the Gumstix are DTE devices.
%The serial port configuration is:

@@

DSP Compoments are deployed as OSGi bundles. OSGi defines bundles life-cycle and
runtime upgrade mechanism. DSP uses these capabilities to allow remote deployment 
(bundles with DSP Components are transmitted as message to the DSP Component Adaptor) 
and activation/deactivation of DSP Components.

Our implementation of the DSP is based on the Open-Source
implementation of OSGi called Knopflerfish \cite{knopflerfish01}.
Since Sun Microsystems does not support its JDK for the Gumstix, we
use JamVM \cite{jamvm01} as the Java virtual machine implementation
and the GNU Classpath \cite{classpath01} project for the runtime libraries. In
the following sections, we provide further details of some of the DSP
Components that we have implemented as part of the NetBEAMS project.

@@

One of the NetBEAMS goals is to easily configure the sensor
network without programming skills. Technicians should be able to
change configuration parameters of various DSP Components via an
easy-to-use web interface. With regards to the NetBEAMS specific
components discussed in earlier sections, these configuration
parameters are, for example, the sampling frequency of the YSI Sonde DP
and the upload frequency of the the wire transport DC.

This configuration is performed by a special DSP Component that
offers a web interface through which a technician can define the
configuration parameters (component \ding{197} in Figure
\ref{FIG_NETBEAMS}). The web management component first queries the remote
component to get a list of configuration parameters. Once the remote
component sends a reply, a user interface is automatically built up
to allow the entry of user defined values. When those values are
submitted, an update message is sent back to the remote component that
will then change the appropriate configuration parameters.

@@

%The California coastal region is populated by a vast number of
%remotely deployed sensors that monitor environmental conditions
%ranging from weather to water quality to ocean surface currents.
%These sensors are operated by data providers, including resource
%management groups, research scientists, municipalities, state and
%federal agencies, who could use near real-time data streams to inform
%decisions that impact public health, public safety, and the protection
%of resources. Success of these efforts, whether small scale or a large
%integrated network, depends on the first step in the data management
%process: reliably obtaining data in near real-time from remotely
%located sensors.  This step can be the weak link in the data
%management pathway when telemetry options are unreliable.  The
%availability of a plug-and-play computing platform designed to
%communicate with a variety of sensor types would stabilize data
%streams with poor telemetry and allowing expansion of sensor coverage
%into regions not presently served.

In this paper, we introduce a practical application of a sensor network
in the domain of environmental monitoring. We have built an
environmental sensor network in the San Francisco Bay using highly
specialized, off-the-shelf water quality sensors. The resulting
infrastructure enables near-realtime access to sensor data collected
from remote locations. Our end-to-end solution, furthermore, allows
configuration and management of remotely deployed sensors. In this sense, the
key to our infrastructure is the DSP, a micro-kernel plug-and-play
architecture that can flexibly be configured to specific needs through
special purpose components. We plan to use the Gumstix's processing
power to pre-process data intensive sensors such as the HF Radar to
reduce the data volume.

Furthermore, our work improves upon existing technologies in earlier sensor network implementations.
Many sensor networks use PIC microcontrollers \cite{Pic01} and specialized operating systems
such as TinyOS \cite{tinyos01}. However, PICs are usually too buggy,
requiring non-standard operating systems \cite{Martinez04}. Likewise, TinyOS exhibits limited 
constraints on storage, restricting any complex processing of sensor data
that can be performed \cite{Martinez04}. OSGi has been used in other sensor network
projects \cite{Mueller07, Baude07}. Although some OSGi sensor networks also include remote 
communication capabilities \cite{Rellermeyer}, our DSP is designed to support asynchronous 
communication in loosely coupled networks. The NetBEAMS infrastructure, as
described in the paper, is available as an Open-Source project and can be 
viewed in operation at \texttt{www.netbeams.org}.

>>>>

\author{\authorblockN{A. Puder, K. Sales, M. de Sales, T. Johnson}
\authorblockA{San Francisco State University \\
Computer Science Department \\
1600 Holloway Avenue \\
San Francisco, CA 94132 \\
EMail: \{arno$\mid$kleber$\mid$msales$\mid$tlj\}@sfsu.edu}
\and
\authorblockN{D. Robinson}
\authorblockA{San Francisco State University \\
Romberg Tiburon Center \\
3150 Paradise Drive \\
Tiburon, CA 94920 \\
EMail: dhr@sfsu.edu}}

\maketitle
\begin{abstract}
 In this paper, we describe a sensor network for environmental
 monitoring based on highly specialized, off-the-shelf sensing
 devices. Environmental monitoring depends on the reliable
 collection of measurements from remotely deployed sensors and the
 rapid transfer of those measurements to data centers. Typically,
 sensors are deployed in out-of-the-way locations, where physical
 access is limited and connections for telemetry are poor or
 nonexistent. Maintaining the data stream from these sensors places
 demands on human resources, requiring site visits to service the
 sensors and to download data stored on internal memory. The work
 described in this paper is a collaborative effort between the
 Computer Science Department at San Francisco State University and
 the Romberg Tiburon Center. The result of this collaboration is an
 environmental sensor network deployed in the San Francisco Bay. It
 provides means to program and interrogate sensors at the remote
 field locations and to transmit collected measurements to a remote
 database.
\end{abstract}

@@

The work described in this paper is a collaborative effort between the
Computer Science Department at San Francisco State University and the
Romberg Tiburon Center (RTC), a research institute focusing on the
understanding of complex marine and estuarine environments. The result
of this collaboration is NetBEAMS (Networked Bay Environmental
Assessment Monitoring System), an environmental sensor network
deployed in the San Francisco Bay. NetBEAMS provides the means to program
and interrogate sensors at remote field locations and to transmit
collected measurements to a database, where the data is rapidly
processed, archived, and made available to potential users in near
realtime. In the following sections we give an overview of the NetBEAMS
architecture and implementation. Section \ref{SEC_BACKGROUND}
describes a typical environmental sensor used by the RTC. Section
\ref{SEC_DSP} describes our Data Sensor Platform (DSP) that allows us
to efficiently build up an end-to-end environmental sensor network while
Section \ref{SEC_NETBEAMS} describes NetBEAMS. Finally, Section 
\ref{SEC_CONCLUSION} presents our conclusions and future works.

@@

Environmental researchers typically use standard, off-the-shelf
sensing devices for their purposes. These sensing devices are at the
opposite end of the size spectrum of miniature sensors such as SmartDust. One such
device is the YSI 6600EDS V2 Sonde \cite{Sonde01}, a water
quality monitoring device that gathers water quality data, and
operates in fresh, sea, or polluted water. Some of the measurements
that the YSI is capable of are conductivity, temperature, chloride,
among others. It is under 55 cm in length, 8.9 cm in diameter, and weighs
approximately 3.18 kg, and operates at temperatures between -5 to +50 deg C and
at depths up to 200 meters. It uses 8 C-size Alkaline Batteries or External 12 VDC,
making the device last up to 75 days depending on the sensor configuration.
The Sonde's bulkhead contains multiple pin connectors to support
sensor probes for measuring parameters as shown in Table
\ref{TAB_SONDE_MEASUREMENTS}.

@@

Each probe may have multiple sensors. A computer can interface with
the Sonde via RS-232C or SDI-12. The Sonde can be configured to
collect data samples in discrete or unnattended mode. Discrete mode
is normally performed while a technician manages the Sonde during the
sampling process. In this mode, the Sonde is connected via 650
MDS Display/Logger or a serial cable connected to a PC, where the sampling
frequency is usually set to a high frequency. In contrast, unattended
mode is performed when the Sonde is deployed offshore or in a
remote body of water, where the sampling frequency is usually set for a
longer period of time (i.e. 5 - 15 min). Assuming a 5 min sampling
frequency, the Sonde produces 267 KB of data per day.
In order to be configured,
the Sonde offers a set of menus that can be accessed through a terminal session
over RS232. In this way, the Sonde can be configured to
start logging data to an internal file. In order to view the data as it is
being captured, the option to show live data is chosen and the data can then be
seen via a serial terminal session at each interval.

We provide these details to show that sensing devices such as the YSI
are highly specialized devices with a unit price of several thousand
dollars. These sensors are usually sparsely deployed in remote
locations that not only make their maintenance difficult, but also
access to their measurements difficult. Typically, the sensors are autonomous
in the sense that they can run for a certain amount of time on battery power and
store the measurement data in internal memory. Only when the sensors
are serviced, the measurement data, which was collected since the last
servicing, is uploaded to a portable storage device. From there, the
data can be uploaded to server. This manual procedure to retrieve
sensor data is not only error prone, but the results greatly differ
between the time the sensor performs the measurement and when
it becomes available for further processing. For many applications,
however, it would be beneficial to have near-realtime access to the
sensor measurements. In this way, we identify the following requirements
that an end-to-end infrastructure supporting devices such as the YSI should
offer:

@@

Sensor networks, especially environmental sensors, are normally
deployed over a geographically wide area, where the communication
infrastructure is not always reliable or, at least, not continuously
available. As the sensor network grows in size and complexity,
scalability, reliability and maintainability become critical issues.
For instance, additional components might be added to the network, sensor
configuration might be altered, or new versions of components might have to be
deployed. As a consequence, the sensor network must be able to support such
adjustments without significant overhead for maintainers. From a system
development and integration point of view, interfaces among components
and between components and the network should be simple. As a result, complex
interaction can be constructed on top of these simple interfaces. In
the following subsections, we discuss how our proposed architecture addresses
those issues.

@@

NetBEAMS's architecture is based on a micro-kernel approach. The core
functionality is limited to a minimum and all extra services are built
as plug-and-play modules on top of the micro-kernel. With the help of
this approach, we can keep the basic infrastructure compact while
allowing customization via special-purpose modules. In our terminology,
we call this basic foundation the \emph{Data Sensor Platform}
(DSP). DSP is a collection of local built-in and plug-in components;
a sensor network is built from several DSPs that are linked
with each other.

Figure \ref{FIG_DSP} depicts the basic architecture of the DSP. We
refer to the plug-and-play modules as DSP Components. DSP Components
are self-contained modules that can be added and removed to a DSP
installation at run-time. The basic communication unit within a DSP
is a message. DSP Components can exchange messages of different types.
Actually, all interactions between DSP Components are implemented
through messages yielding a simple, yet powerful communication
structure. Moreover, we distinguish between \emph{Data Producers} (DP) and
\emph{Data Consumers} (DC) within the scope of a DSP. DP and DC can be
seen as roles that a DSP Component may have, as it may also 
act in the role of DC and DP at the same time. Messages
in DSPs are routed based on rules, which can be configured statically
or dynamically. Rules for routing messages can be based on different
information such as producers or consumers of the message, the message
itself, or specific configuration parameters provided externally.

@@

@@

A DSP Component encapsulates a particular functionality. We
distinguish between those components that produce data (DP) and those
that consume data (DC). An example for a DSP Component acting in the
role of a DP is a module that interfaces with a physical sensor. Such
a module reads data from the sensor and converts the data into a
message that is forwarded to the DSP. Because this module generates
data with respect to the DSP, it is a data producer. An example for a
data consumer is a converter module that converts the internal,
DSP-specific data format to a format suitable for consumption by
external applications. One such common format used in the domain of
environmental monitoring is OPEnDAP \cite{opendap01}. Since the
converter module accepts messages from the DSP, it acts in the role of
a data consumer. Finally, a DSP Component implementing persistence is
an example of a module that acts in the roles of both data producer
and consumer, since it accepts messages from the DSP and stores them in a
database. Furthermore, it may act as a data producer by resending previously
stored messages at a later time.

@@

By implementing the above Java interface, the resulting code
effectively becomes a DSP Component that can be deployed on a DSP at
run-time. A component is identified by its type (line 3). Only one
component of a certain type per DSP is permissible. A type identifier
must be unique across all DSP Components in a DSP. Upon deploying a
component, an initialization function is called (line 4). Once a
component is initialized, it may be started (line 6) and stopped (line
7) multiple times. Whenever the DSP wishes to pass a message (that was
generated by some DP), it invokes the \texttt{deliver()} method (line
8). A DP can send messages via a \texttt{DSP Broker} through the 
\texttt{DSPContext}, which is passed to the DP during initialization.

@@

The broker can deliver messages concurrently and asynchronously. It
sets any additional message header information necessary for proper
routing. The broker does not make any routing decisions. It delegates
this responsibility to the \emph{matcher}. The matcher is responsible
for matching messages sent by a DP and routes it to one or more DCs.
This decision is based on a set of routing rules (see Figure
\ref{FIG_MATCHER_RULE}), which can be statically or dynamically
configured.

@@

The matcher's main purpose is to achieve a loose coupling between
DCs and DPs. DPs may create messages without specifying a specific
destination DC. Once the message is sent through the DSP, the matcher will
determine which DCs should receive the message. Note that a DP can
still specify a DC for its message. In this case, the matcher includes
this DC in a list of DCs to receive the message.

@@

The Gumstix runs Linux 2.6 with the BusyBox utilities \cite{busybox01}, and uses
the OpenEmbedded cross-compile build environment to provide a complete
Linux environment and a large range of Linux applications for the ARM
architecture. We use the JamVM \cite{jamvm01} implementation of Sun
Microsystem's virtual machine in order to run the DSP, together with
the GNU Classpath \cite{classpath01} as the runtime libraries. 
The Gumstix communicates with the YSI Sonde via an RS232 serial port using a
null-modem cable, since both the Sonde and the Gumstix are DTE devices.
%The serial port configuration is:

@@

DSP Compoments are deployed as OSGi bundles. OSGi defines bundles life-cycle and
run-time upgrade mechanism. In this way, DSP uses these capabilities to allow remote deployment 
(bundles with DSP Components are transmitted as message to the DSP Component Adaptor) 
and activation/deactivation of DSP Components. Our solution is implemented using 
Knopflerfish \cite{knopflerfish01}, an Open-Source OSGi implementation.

In the following sections, we provide further details of some of the DSP
Components that we have implemented as part of the NetBEAMS project.

@@

One of the NetBEAMS goals is to easily configure the sensor
network without programming skills. Technicians should be able to
change configuration parameters of various DSP Components via an
easy-to-use web interface. As discussed in earlier sections, these 
configuration parameters are, for example, the sampling frequency of the YSI
Sonde DP or the upload frequency of the the wire transport DC.

This configuration is performed by a special DSP Component that
offers a web interface through which a technician can define
configuration parameters (component \ding{197} in Figure
\ref{FIG_NETBEAMS}). The web management component first queries the remote
component to get a list of configuration parameters. Once the remote
component sends a reply, a user interface is automatically built up
to allow the entry of user defined values. When those values are
submitted, an update message is sent back to the remote component to be 
adjusted accordingly.

@@

%The California coastal region is populated by a vast number of
%remotely deployed sensors that monitor environmental conditions
%ranging from weather to water quality to ocean surface currents.
%These sensors are operated by data providers, including resource
%management groups, research scientists, municipalities, state and
%federal agencies, who could use near realtime data streams to inform
%decisions that impact public health, public safety, and the protection
%of resources. Success of these efforts, whether small scale or a large
%integrated network, depends on the first step in the data management
%process: reliably obtaining data in near realtime from remotely
%located sensors. This step can be the weak link in the data
%management pathway when telemetry options are unreliable. The
%availability of a plug-and-play computing platform designed to
%communicate with a variety of sensor types would stabilize data
%streams with poor telemetry and allowing expansion of sensor coverage
%into regions not presently served.

In this paper, we introduce a practical application of a sensor network
in the domain of environmental monitoring. We have built an
environmental sensor network in the San Francisco Bay using highly
specialized, off-the-shelf water quality sensors. The resulting
infrastructure enables near-realtime access to sensor data collected
from remote locations. Our end-to-end solution, furthermore, allows
configuration and management of remotely deployed sensors. In this sense, the
key to our infrastructure is the DSP, a micro-kernel plug-and-play
architecture that can be configured to specific needs through
special purpose components. We plan to use the Gumstix's processing
power to pre-process data intensive sensors such as the HF Radar to
reduce the data volume.

Furthermore, our work improves upon existing technologies in earlier sensor network implementations.
Many sensor networks use PIC microcontrollers \cite{Pic01} and specialized operating systems
such as TinyOS \cite{tinyos01}. However, PICs are usually too buggy,
requiring non-standard operating systems \cite{Martinez04}. Likewise, TinyOS exhibits limited 
constraints on storage, restricting any complex processing of sensor data
that can be performed \cite{Martinez04}. OSGi has been used in other sensor network
projects \cite{Mueller07, Baude07}. Although some OSGi sensor networks also include remote 
communication capabilities \cite{Rellermeyer}, our DSP is designed to support asynchronous 
communication in loosely coupled networks. The NetBEAMS infrastructure is available as an Open-Source
project and can be viewed in operation at \texttt{www.netbeams.org}.
